---
title: "Enrichment Analysis of H3K27ac peaks III"
output: html_document
---

```{r simulation_options, include=FALSE}
library(knitr)
library(rmarkdown)
library(report)
library(stringr)
library(ChIPseeker)
library(DESeq2)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)
library(RColorBrewer)
library(dplyr)
library(ggplot2)
library(reshape2)
library(GenomicFeatures)
opts_chunk$set(fig.width = 10, fig.height = 10, dpi = 72,
               warning = FALSE, message = FALSE, prompt = TRUE, cache = FALSE)
set.seed(123)
```
last changed: `r Sys.Date()`

---

Feature annotation of H3K27ac peaks
===

## Data processing

Load the `rlog` transformed DESeq data set and construct the required data 
frames. The function `assay` returns the values and `colData` returns the 
information about the experimental design.
```{r data1}
rlog.heat <- readRDS(file = "./../Significance_testing/Data/rlog_H3K27ac_sig.rds")
head(rlog.heat)

m.H3.ac <- assay(rlog.heat)
head(m.H3.ac)

info.H3.ac <- data.frame(colData(rlog.heat))
paged_table(info.H3.ac)
```

---

Next the `GRanges` object can be constructed using the `rownames` of the DESeq 
data set, which contain information about the location of each peak.
```{r GR}
chr <- str_split_i(rownames(m.H3.ac), ":", 1)
pos <- str_split_i(rownames(m.H3.ac), ":", 2)
start <- as.integer(str_split_i(pos, "-", 1))
end <- as.integer(str_split_i(rownames(m.H3.ac), "-", 2))

peaks.H3ac <- GRanges(seqnames = chr, 
                        ranges =IRanges(start = start, end = end))
head(peaks.H3ac)

saveRDS(peaks.H3ac, "./Data/peaks_H3ac.rds")
```

---

To visualize the distance to TSS the by cluster, a `list` containing a entry for  
each cluster with the corresponding peaks is required. Thus, the clustering 
results are loaded and used to filter the `GRanges` object.
```{r clus_4}
kmeans.k3 <- readRDS("./../Clustering/Data/kmeans.rds")

lapply(kmeans.k3, head)

clus.kmeans <- melt(kmeans.k3$cluster)
clus.kmeans <- cbind(clus.kmeans, 1:nrow(clus.kmeans))
colnames(clus.kmeans) <- c("cluster", "ind")
clus.kmeans <- clus.kmeans[order(clus.kmeans[, "cluster"]), ]

paged_table(clus.kmeans)

ind.cluster1 <- clus.kmeans[which(clus.kmeans[, "cluster"] == 1), "ind"]
ind.cluster2 <- clus.kmeans[which(clus.kmeans[, "cluster"] == 2), "ind"]
ind.cluster3 <- clus.kmeans[which(clus.kmeans[, "cluster"] == 3), "ind"]

peaks.k1 <- peaks.H3ac[ind.cluster1, ]
peaks.k2 <- peaks.H3ac[ind.cluster2, ]
peaks.k3 <- peaks.H3ac[ind.cluster3, ]

peaks <- list(cluster_1 = peaks.k1, cluster_2 = peaks.k2, 
              cluster_3 = peaks.k3)
```

---

## Average distance to TSS

The `GRanges` list can then be used to plot the average distance to the TSS by 
cluster.
```{r AvgProf}
hg38 <- TxDb.Hsapiens.UCSC.hg38.knownGene
promoter <- getPromoters(TxDb = hg38, upstream=3000, downstream=3000)
tagM.peaks <- lapply(peaks, getTagMatrix, windows=promoter)

plotAvgProf(tagM.peaks, TxDb = hg38, xlim=c(-3000, 3000), 
             xlab = "Genomic Region (5'->3')", ylab = "Peak Count") +
  scale_color_manual(values = brewer.pal(n = 3, name = "Dark2"))
```

---

## Feature annotation

Next, the peaks were annotated to genes using `annotatePeak` and all known human 
genes. The resulting object was then plotted with `plotAnnoBar`, to visualize the 
distribution of features among peaks.
```{r anno}
anno.list <- lapply(peaks, annotatePeak, TxDb = hg38, 
                         tssRegion=c(-3000, 3000), verbose=FALSE)

plotAnnoBar(anno.list)
```

---

Enrichment Analysis
===

## Data processing

For the enrichment analysis the peaks were annotated to only coding genes within 
a range of -1 to 1 kb from the transcription start site (TSS). For this the 
peaks are annotated to using only the protein coding genes.
```{r anno.cod}
TxDb.protein.coding <- loadDb( file="./Data/TxDb.protein.coding.sqlite")
seqlevelsStyle(TxDb.protein.coding) <- "UCSC"

peak.coding <- lapply(peaks, annotatePeak, TxDb = TxDb.protein.coding, 
                      tssRegion = c(-3000, 3000), verbose=FALSE)
```

---

Next, the following function was used to translate the ENSEMBL identifiers to 
ENTREZID.
```{r ID}
# generate required structure:
conv2table<-function(x){
  eg = bitr(x@anno$geneId, fromType="ENSEMBL", toType="SYMBOL", OrgDb="org.Hs.eg.db")
  ez = bitr(x@anno$geneId, fromType="ENSEMBL", toType="ENTREZID", OrgDb="org.Hs.eg.db")
  anno.table<-x@anno
  #add gene symbols
  mx<-match(anno.table$geneId, eg$ENSEMBL)
  anno.table$gene_symbol<-eg[mx,"SYMBOL"]
  #add gene entrezID
  mx<-match(anno.table$geneId, ez$ENSEMBL)
  anno.table$gene_entrez<-ez[mx,"ENTREZID"]
  
  res.table<-cbind(
    chr=as.vector(seqnames(anno.table)),
    start=start(anno.table),
    end=end(anno.table),
    name=anno.table$name,
    peak.p_adj=anno.table$qValue,
    strand=rep(".", length(anno.table)),
    genome_anno=anno.table$annotation,
    gene_symbol=anno.table$gene_symbol,
    gene_entrez=anno.table$gene_entrez,
    distanceToTSS=anno.table$distanceToTSS,
    transcriptId=anno.table$transcriptId,
    geneId_ensembl=anno.table$geneId)
}

peak.coding.ID <- lapply(peak.coding, function(n) conv2table(n))
lapply(peak.coding.ID, head)
```

---

The resulting list was then filtered for peaks which are within a range of -1 to 
1 kb from the TSS.
```{r heat}
peak.coding.1kb <- lapply(peak.coding.ID, function(x) 
  filter(as.data.frame(x), 
         abs(as.integer(as.data.frame(x)$distanceToTSS)) <= 1000))

lapply(peak.coding.1kb, head)

saveRDS(peak.coding.1kb, file = "./Data/peaks_H3ac_cod1kb.rds")

names(peak.coding.1kb) <- c("1", "2", "3")
cod.genes.1kb <- lapply(peak.coding.1kb, function(i) as.data.frame(i)$gene_entrez)
```

---

## Reactome pathway analysis enrichment analysis

Finally, the filtered list with the coding genes can be used for an enrichment 
analysis using the function `enrichPathway`, which can be plotted using 
`dotplot`.
```{r Reactome, fig.width = 10, fig.height = 18}
comp.Reac <- compareCluster(geneCluster = cod.genes.1kb,
                            fun = "enrichPathway",
                            pvalueCutoff = 0.05,
                            pAdjustMethod = "BH")

comp.Reac %>% as.data.frame() %>% paged_table()

dotplot(comp.Reac, showCategory = 10, 
        title = "Reactome Pathway Enrichment Analysis (coding genes)")
```

---

## Gene Ontology enrichment analysis

Exactly the same as above can be done with the function `enrichGO`.
```{r GO, fig.width = 10, fig.height = 18}
comp.GO <- compareCluster(geneCluster = cod.genes.1kb,
                          fun = "enrichGO",
                          OrgDb = org.Hs.eg.db,
                          ont = "BP",
                          readable = T,
                          pvalueCutoff = 0.05,
                          pAdjustMethod = "BH")

comp.GO %>% as.data.frame() %>% paged_table()

dotplot(comp.GO, showCategory = 10, 
        title = "GO Enrichment Analysis (coding genes)")
```
---

References - packages
---------------------

```{r packages, echo=FALSE, results='asis'}
cite_packages()
```

---

Session Information
-------------------

The following versions of R and R packages were used to generate the report
above:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```