---
title: "Significance testing for H3K27ac peaks II"
output: html_document
---

```{r simulation_options, include=FALSE}
library(knitr)
library(rmarkdown)
library(report)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(reshape2)
opts_chunk$set(dpi = 72,
               warning = FALSE, message = FALSE, prompt = TRUE, cache = FALSE)
set.seed(123)
```
last changed: `r Sys.Date()`

---

## Significance testing (Wald)

Since the time points 0 hpi and 2 hpi showed the most difference in the PCA (see:
LRT/Significance_testing), the corresponding samples were filtered out. For this 
the same peaks are loaded as generated in the significance testing for LRT. 
After filtering a new DESeq data set was generated, only containing the samples 
for 0 and 2 hpi.
```{r data1}
rlog.H3ac <- readRDS(file = "./../../1_LRT/Significance_testing/Data/rlog_H3K27ac.rds")
head(rlog.H3ac)

H3ac.raw <- readRDS(file = "./../../1_LRT/Significance_testing/Data/raw_H3K27ac.rds")
head(H3ac.raw)

m.raw <- assay(H3ac.raw)
head(m.raw)

info.raw <- data.frame(colData(H3ac.raw))
paged_table(info.raw)

info.raw.02hpi <- info.raw[which(info.raw$time == 0 | info.raw$time == 2), ]

m.raw.02hpi <- m.raw[, rownames(info.raw.02hpi)]

dds.02hpi <- DESeqDataSetFromMatrix(m.raw.02hpi, 
                                    info.raw.02hpi, 
                                    design = ~ construct + replica + time)
head(dds.02hpi)
```

---

To find significant peaks a Wald pairwise significance test was used. The output 
was then used to filter the rlog data for significant peaks (adjusted p value 
<= 0.05).
```{r Wald}
dds.Wald <- DESeq(dds.02hpi, test = "Wald")
head(dds.Wald)

res <- results(dds.Wald)
head(res)

res.sig <- res[which(res$padj <= 0.05), ]

rlog.H3ac.sig <- rlog.H3ac[rownames(res.sig), ]

saveRDS(rlog.H3ac.sig, file = "./Data/rlog_H3K27ac_sig.rds")
```

---

References - packages
---------------------

```{r packages, echo=FALSE, results='asis'}
cite_packages()
```

---

Session Information
-------------------

The following versions of R and R packages were used to generate the report
above:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```