---
title: "Significance testing for H3K27ac peaks"
output: html_document
---

```{r simulation_options, include=FALSE}
library(knitr)
library(rmarkdown)
library(report)
library(IRanges)
library(dplyr)
library(stringr)
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(pheatmap)
opts_chunk$set(dpi = 72,
               warning = FALSE, message = FALSE, prompt = TRUE, cache = FALSE)
set.seed(123)
```
last changed: `r Sys.Date()`

---

Sample coverage
===

## Generate DESeq data set

First, the new coverage file is loaded using `read.table`. Then, the columns 
which contain the information about the peak position are combined into one 
vector (`seq.names`). Next, only the values of the coverage file are used and
the peak position is annotated as row names.
```{r data1}
cov.H3ac <- read.table("./Data/H3K27ac_coverage.tabular", 
                       header = T)

head(cov.H3ac)

seq.names <- apply(cov.H3ac[, 1:3], 1, function(row) {paste(row[1], 
                                                         ":", 
                                                         row[2] %>% as.integer(), 
                                                         "-", 
                                                         row[3] %>% as.integer(), 
                                                         sep = "")})

m.H3ac <- cov.H3ac[ , -which(names(cov.H3ac) %in% c("chr", "start", "end"))]
rownames(m.H3ac) <- seq.names
colnames(m.H3ac) <- str_split_i(colnames(m.H3ac), ".bam", 1)

head(m.H3ac)
```

---

For generating a DESeq data set a `colData` file is required containing 
information about the experiment. Note that the parameter `time` is required to 
be `numeric` which is essential for a meaningful significance testing. Thus, the 
required information were gathered in a matrix, as follows:
```{r colData}
colData.H3ac <- matrix(NA, nrow = 22, ncol = 3) %>% as.data.frame()
rownames(colData.H3ac) <- colnames(m.H3ac)
colnames(colData.H3ac) <- c("construct", "replica", "time")
colData.H3ac$construct <- str_split_i(colnames(m.H3ac), "_", 1) %>% 
  as.factor()
colData.H3ac$replica <- str_split_i(colnames(m.H3ac), "_", 3) %>% 
  as.factor()
colData.H3ac$time <- str_split_i(colnames(m.H3ac), "T", 2) %>% 
  str_split_i( "h", 1) %>% 
  as.numeric()

colData.H3ac <- arrange(colData.H3ac, construct, time, replica)

paged_table(colData.H3ac)

m.H3ac <- m.H3ac[ , rownames(colData.H3ac)]
```

---

Now that we have all required files, the DESeq data set can be constructed and 
the parameter `design` takes the parameters for the underlying model.
```{r clus_3}
H3ac.coverage <- DESeqDataSetFromMatrix(m.H3ac, 
                                        colData.H3ac, 
                                        design = ~ construct + replica + time)

saveRDS(H3ac.coverage, file = "./Data/raw_H3K27ac.rds")

head(H3ac.coverage)
```

---

## Regularized-log transformation and PCA

Next, the data was scaled using the `rlog` transformation to stabilize the data 
against variance within the data.
```{r rlog}
rlog.H3ac <- rlog(H3ac.coverage, blind = F)

saveRDS(rlog.H3ac, file = "./Data/rlog_H3K27ac.rds")

head(rlog.H3ac)
```

---

To validate the data the rlog-transformed data was used to plot a principal 
component analysis (PCA) and a heatmap with hierachial clustering.
```{r PCA, fig.width = 12, fig.height = 10}
pca.H3ac <- plotPCA(rlog.H3ac, 
                    intgroup = c("construct", "time"), 
                    returnData = T)
per.Var <- round(100 * attr(pca.H3ac, "percentVar"))

ggplot(pca.H3ac, aes(PC1, PC2, color = time %>% as.factor(), shape = construct, label = name)) +
  geom_point(size = 3) +
  geom_text_repel(size = 2.5) +
  labs(x = paste0("PC1: ", per.Var[1], "% variance"), 
       y = paste0("PC2: ", per.Var[2], "% variance"),
       color = "time",
       main = "PCA for H3K27ac after infection in hours") +
  theme_bw()
```

```{r heat, fig.width = 12, fig.height = 10}
sample.dist <- dist(t(assay(rlog.H3ac)))
m.dist <- sample.dist %>% as.matrix()
colnames(m.dist) <- NULL

col <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

plot.heat <- pheatmap(m.dist, 
                      clustering_distance_rows = sample.dist, 
                      clustering_distance_cols = sample.dist, 
                      col = col)
```

---

## Significance testing (Likelihood ratio test - LRT)

To find significant peaks a likelihood ratio test (LRT) was used. This test 
compares 2 models, the full model (M ~ construct + replica + time) and the 
simpler model (M(s) ~ construct + replica). This results in peaks which show 
significant difference between those 2 models, in this case only peaks which 
significantly affect the larger model will score a low enough p-values. The 
output was then used to filter the rlog data for significant peaks (adjusted p 
value <= 0.05).
```{r LRT}
H3ac.LRT <- DESeq(H3ac.coverage, test = "LRT", reduced = ~ construct + replica)

head(H3ac.LRT)

res <- results(H3ac.LRT)

head(res)

res.sig <- res[which(res$padj <= 0.05), ]

rlog.H3ac.sig <- rlog.H3ac[rownames(res.sig), ]

saveRDS(rlog.H3ac.sig, file = "./Data/rlog_H3K27ac_sig.rds")
```

---

References - packages
---------------------

```{r packages, echo=FALSE, results='asis'}
cite_packages()
```

---

Session Information
-------------------

The following versions of R and R packages were used to generate the report
above:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```