---
title: "Clustering of H3K27ac peaks III"
output: html_document
---

```{r simulation_options, include=FALSE}
library(knitr)
library(rmarkdown)
library(report)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(reshape2)
opts_chunk$set(fig.width = 10, fig.height = 10, dpi = 72,
               warning = FALSE, message = FALSE, prompt = TRUE, cache = FALSE)
set.seed(123)
```
last changed: `r Sys.Date()`

---

Hierarchical Clustering
===

Load the `rlog` transformed DESeq data set and construct the required data 
frames. The function `assay` returns the values and `colData` returns the 
information about the experimental design.
```{r data1}
rlog.heat <- readRDS(file = "./../Significance_testing/Data/rlog_H3K27ac_sig.rds")
head(rlog.heat)

m.H3.ac <- assay(rlog.heat)
head(m.H3.ac)

info.H3.ac <- data.frame(colData(rlog.heat))
paged_table(info.H3.ac)
```

---

Next this data can be used to perform hierarchical clustering, which is 
implemented in `pheatmap`. `colorRampPallette` was used to change the annotation 
color.
```{r hier_clus}
col <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
ann.col <- list(construct = c(U2OS31 = "#FFCC33", U2OS33 = "#CC3366"))

heat.hieracial <- pheatmap(m.H3.ac, 
                  cluster_cols = F, 
                  show_rownames = F, 
                  scale = "row", 
                  color = col,
                  annotation_col = info.H3.ac[, c("construct", "time")],
                  annotation_colors = ann.col)
```

---

Kmeans Clustering
===

To get a better separation of the data, kmeans clustering was performed. Since
the H3K27ac data was yielded from two separate constructs U2OS31 (histone H3.1) 
and U2OS33 (histone H3.3), the data was separated by construct and scaled to 
Z-scores using the function `scale`.
```{r data2}
m.H3.1 <- m.H3.ac[, info.H3.ac$construct == "U2OS31"]
m.H3.3 <- m.H3.ac[, info.H3.ac$construct == "U2OS33"]

norm.m.H3.1 <- t(scale(t(m.H3.1)))
norm.m.H3.3 <- t(scale(t(m.H3.3)))

norm.m.H3 <- cbind(norm.m.H3.1, norm.m.H3.3)
```

---

Next, `kmeans` was used to perform the clustering and a factor indicating the 
clusters was added to each of the scaled matrices. The clustering for
`centers = 3` (3 cluster) yielded the best result, as more clusters divided 
data with a similar course.
```{r kmeans}
clus.kmeans <- kmeans(norm.m.H3, 3)

lapply(clus.kmeans, head)

saveRDS(clus.kmeans, "./Data/kmeans.rds")

m.kmeans <- cbind(norm.m.H3, cluster = as.factor(clus.kmeans$cluster))
m.kmeans <- m.kmeans[order(m.kmeans[, "cluster"]), ]

m.kmeans.H31 <- m.kmeans[, c(1:10, 23)]
m.kmeans.H33 <- m.kmeans[, 11:23]
```

---

To visualize the clustering a heatmap was plotted using `pheatmap`. The functions 
`colorRampPalette` and `brewer.pal` were used to generate the color for the 
annotations and Z-values, respectively.
```{r heat, fig}
col <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
col2 <- brewer.pal(n = 8, name = "Dark2")
clus.col <- c("1" = col2[1], "2" = col2[2], "3" = col2[3])
ann.col2 <- list(construct = c(U2OS31 = "#FFCC33", U2OS33 = "#CC3366"), cluster = clus.col[1:3])

heat.kmeans.H31 <- pheatmap(m.kmeans.H31[ ,-ncol(m.kmeans.H31)],
                        cluster_rows = F,
                        cluster_cols = F,
                        show_rownames = F, 
                        color = col,
                        annotation_col = info.H3.ac[, c("construct", "time")],
                        annotation_row = data.frame(cluster = as.factor(m.kmeans[ ,"cluster"])),
                        annotation_colors = ann.col2)
  
heat.kmeans.H33 <- pheatmap(m.kmeans.H33[ ,-ncol(m.kmeans.H33)],
                        cluster_rows = F,
                        cluster_cols = F,
                        show_rownames = F, 
                        color = col,
                        annotation_col = info.H3.ac[, c("construct", "time")],
                        annotation_row = data.frame(cluster = as.factor(m.kmeans[, "cluster"])),
                        annotation_colors = ann.col2)
```

---

Lastly, the cluster centroids (Z-scores) were plotted separately to visualize 
the change in H3K27ac over time between the two constructs.
```{r Z_cen}
kmeans <- melt(clus.kmeans$centers)
ind <- match(kmeans$Var2, rownames(info.H3.ac))
data.kmeans <- cbind(kmeans, info.H3.ac[ind, c("time", "construct")])
colnames(data.kmeans) <- c("cluster", "name", "Z_score", "time", "construct")

paged_table(data.kmeans)
  
ggplot(data.kmeans) + 
  geom_point(aes(x = time, y = Z_score, color = construct), size = 1) +
  stat_summary(aes(x = time, y = Z_score, color = construct), fun = mean, geom = "line", linewidth = 1) +
  facet_grid(rows = data.kmeans$cluster) +
  scale_color_manual(values = c("#FFCC33", "#CC3366")) +
  theme_bw() +
  labs(title = "z-scores for H3K27ac after infection in hours", y = "z-scores")
```

---

References - packages
---------------------

```{r packages, echo=FALSE, results='asis'}
cite_packages()
```

---

Session Information
-------------------

The following versions of R and R packages were used to generate the report
above:

```{r sessionInfo, echo=FALSE}
sessionInfo()
```